---
title: "What Are Youth Doing in Economies of the Arab World?"
subtitle: "Exploring Arab Barometer Wave VIII Data"
author: "R.Kam"
institute: "American University"
date: "May 6, 2026"
format:
  beamer:
    theme: Madrid
    colortheme: default
    fonttheme: structurebold
    slide-level: 2
    toc: false
    incremental: false
    keep-tex: false
    fig-width: 5.7
    fig-height: 2.8
classoption: "aspectratio=169"
---

## Why This Matters

**The personal question:**

I'm graduating soon. Job market feels uncertain. Economy feels shaky. So I wondered: what are young people doing when their economies fail them?

\medskip

**What I found:**

Arab Barometer surveys 8 countries experiencing economic stress. Their youth (ages 18-35) face a choice: **exit** (leave) or **voice** (protest)?

\medskip

**The puzzle:**

Some countries see mass emigration plans. Others see protests. Some see neither. **Why the difference?**

---

## The Data Journey

```{r setup}
#| echo: false
#| warning: false
#| message: false

# The tools I need for this analysis
library(tidyverse)      # For data wrangling
library(knitr)          # For making nice visuals/tables
library(kableExtra)     # For making really nice & fancy tables
library(broom)          # For tidying regression outputs
library(survey)         # For handling weighted survey data
library(ggrepel)        # For non-overlapping labels in plots
```

```{r data-load}
#| echo: false
#| warning: false
#| message: false

# Load the raw survey data
ab8 <- read_csv("ArabBarometer_WaveVIII_English_v3.csv")

# First, I translate country codes into actual names
# The survey uses numbers (7, 8, 9...) but readers need to see "Iraq", "Jordan"...
country_labels <- c(
  "7" = "Iraq", "8" = "Jordan", "9" = "Kuwait", "10" = "Lebanon",
  "12" = "Mauritania", "13" = "Morocco", "15" = "Palestine", "21" = "Tunisia"
)
```

```{r filter-youth}
#| echo: false
#| warning: false
#| message: false

# Add country names and filter to youth (18-35, following World Bank definition)
ab8_youth <- ab8 %>%
  mutate(country_name = factor(COUNTRY, levels = names(country_labels), 
                                labels = country_labels)) %>%
  filter(Q1001 >= 18, Q1001 <= 35)

# How many youth are we studying?
youth_count <- nrow(ab8_youth)
```

**Dataset:** Arab Barometer Wave VIII (Sep 2023 - Jul 2024)

**Sample:** `r format(youth_count, big.mark=",")` youth across 8 countries

**My approach:** Focus on two key behaviors — emigration plans and protest participation

---

## Creating the Key Variables

```{r create-responses}
#| echo: false
#| warning: false
#| message: false

# Here's where I turn raw survey responses into something meaningful
# The survey asks: "Do you plan to emigrate?" (Q501D: 1=Yes, 2=No, 98/99=Refused)
# I recoded this into a simple binary: 1 = planning to leave, 0 = staying
ab8_youth <- ab8_youth %>%
  mutate(emigrate = case_when(
    Q501D == 1 ~ 1,    # Yes, planning to emigrate
    Q501D == 2 ~ 0,    # No, staying
    TRUE ~ NA_real_    # Everything else (refused/don't know) = missing data
  ))

# Same logic for protest: have they EVER participated?
# Q404: 1=Last year, 2=Before, 3=Never
ab8_youth <- ab8_youth %>%
  mutate(protest = case_when(
    Q404 %in% c(1, 2) ~ 1,  # Any participation = 1
    Q404 == 3 ~ 0,          # Never = 0
    TRUE ~ NA_real_         # Refused/don't know = missing
  ))

# I also need to capture economic perceptions and personal circumstances
# Economic assessment: Q101 asks "How is the national economy?" 
# (1=Very good, 2=Good, 3=Bad, 4=Very bad, 5=Neither)
# I reverse it so higher = worse, which makes more intuitive sense
ab8_youth <- ab8_youth %>%
  mutate(
    econ_national = 6 - Q101,  # Now 5 = very bad, 1 = very good
    male = case_when(Q1002 == 1 ~ 1, Q1002 == 2 ~ 0, TRUE ~ NA_real_),
    
    # Wealth: Since Q1015 has limited data, I use Q102 (personal economic situation)
    # Q102: 1=Very good, 2=Good, 3=Bad, 4=Very bad, 5=Neither
    # "Struggling" = those who rate their personal economy as bad/very bad
    wealth_struggling = case_when(
      Q102 >= 98 ~ NA_real_,           # Missing data codes
      Q102 %in% c(1, 2) ~ 0,           # Good/Very good = Comfortable
      Q102 %in% c(3, 4) ~ 1,           # Bad/Very bad = Struggling
      Q102 == 5 ~ NA_real_,            # "Neither" = unclear, set as missing
      TRUE ~ NA_real_
    )
  )
```

```{r country-summary}
#| echo: false
#| warning: false
#| message: false

# Now I calculate country-level summaries
# IMPORTANT: I use survey weights (WT) so results reflect the actual population
ab8_summary <- ab8_youth %>%
  group_by(country_name) %>%
  summarise(
    N = n(),
    emigrate_pct = weighted.mean(emigrate, w = WT, na.rm = TRUE) * 100,
    protest_pct = weighted.mean(protest, w = WT, na.rm = TRUE) * 100,
    econ_bad_pct = weighted.mean(Q101 >= 3, w = WT, na.rm = TRUE) * 100
  )

# Extract key numbers for inline text
top_emigrate_country <- ab8_summary %>% 
  slice_max(emigrate_pct) %>% pull(country_name) %>% as.character()
top_emigrate_pct <- ab8_summary %>% 
  slice_max(emigrate_pct) %>% pull(emigrate_pct) %>% round(1)
```

**What I'm measuring:**

- **Exit:** Planning to emigrate (yes/no)
- **Voice:** Ever participated in protests (yes/no)
- **Context:** How bad is the economy? Personal wealth status?

**Key insight:** Survey weights matter — without them, we'd be counting responses, not representing populations.

---

## The Economic Crisis is Real

```{r fig-economic-crisis}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 5
#| fig.height: 2.6
#| fig.cap: "Widespread economic pessimism across all countries"

# This visualization shows just how bad people think their economies are
# I sort countries by pessimism level to show the gradient
ab8_summary %>%
  ggplot(aes(x = reorder(country_name, econ_bad_pct), y = econ_bad_pct)) +
  geom_col(fill = "firebrick", width = 0.7) +
  geom_text(aes(label = paste0(round(econ_bad_pct, 1), "%")),
            hjust = -0.2, size = 2.2) +
  coord_flip() +
  labs(title = "Economic Pessimism is Widespread",
       subtitle = "Percentage of respondents saying economy is 'bad' or 'very bad'",
       x = NULL, y = "Percentage (%)") +
  scale_y_continuous(limits = c(0, 105), breaks = seq(0, 100, 20)) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(face = "bold", size = 9),
        plot.subtitle = element_text(size = 7),
        panel.grid.major.y = element_blank())
```

\small
**The story:** Lebanon leads at 99% — nearly everyone thinks the economy is terrible. Even Kuwait, the richest, sits at 34%. **Economic stress is universal, but responses differ.**
\normalsize

---

## Two Responses: Exit vs Voice

```{r prepare-comparison}
#| echo: false
#| warning: false
#| message: false

# Now I want to see both responses side-by-side
# I reshape the data from wide (emigrate_pct, protest_pct in separate columns)
# to long format (one column for type, one for percentage)
# This lets ggplot create grouped bars automatically
ab8_comparison <- ab8_summary %>%
  select(country_name, emigrate_pct, protest_pct) %>%
  pivot_longer(cols = c(emigrate_pct, protest_pct),
               names_to = "response_type", 
               values_to = "percentage") %>%
  mutate(response_type = recode(response_type,
                                emigrate_pct = "Immigration",
                                protest_pct = "Protest"))
```

```{r fig-exit-voice}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 5.2
#| fig.height: 2.6
#| fig.cap: "Countries choose different strategies"

# Here I create the side-by-side comparison
# Each country gets two bars: one for exit (orangered), one for voice (blue)
ggplot(ab8_comparison, 
       aes(x = reorder(country_name, percentage), y = percentage, fill = response_type)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = paste0(round(percentage, 1), "%")),
            position = position_dodge(width = 0.7), hjust = -0.2, size = 1.8) +
  coord_flip() +
  scale_fill_manual(values = c("Immigration" = "orangered", "Protest" = "dodgerblue"),
                    name = NULL) +
  labs(title = "Exit vs Voice: Different Countries, Different Choices",
       x = NULL, y = "Percentage (%)") +
  scale_y_continuous(limits = c(0, 65)) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(face = "bold", size = 9),
        legend.position = "bottom",
        legend.text = element_text(size = 7),
        panel.grid.major.y = element_blank())
```

\small
**Three patterns emerge:** Iraq/Mauritania (high protest, lower exit) • Tunisia/Kuwait (high exit, moderate protest) • Lebanon (everyone wants to leave)
\normalsize

---

## Gender Shapes Who Can Leave

```{r gender-analysis}
#| echo: false
#| warning: false
#| message: false

# I suspect gender matters — migration networks are often male-dominated
# Let me calculate emigration rates separately for men and women
gender_data <- ab8_youth %>%
  group_by(country_name, male) %>%
  summarise(emigrate_pct = weighted.mean(emigrate, w = WT, na.rm = TRUE) * 100,
            .groups = "drop")

# Reshape to compare: one row per country, columns for Male/Female/Gap
gender_table <- gender_data %>%
  mutate(gender = ifelse(male == 1, "Male", "Female")) %>%
  select(-male) %>%
  pivot_wider(names_from = gender, values_from = emigrate_pct) %>%
  mutate(gap = Male - Female) %>%
  arrange(desc(gap))

# Pull out the extremes for storytelling
max_gap_country <- gender_table %>% slice_max(gap) %>% pull(country_name) %>% as.character()
max_gap_value <- gender_table %>% slice_max(gap) %>% pull(gap) %>% round(1)
```

\footnotesize

```{r gender-table}
#| echo: false
#| warning: false
#| message: false

kable(gender_table,
      col.names = c("Country", "Female (%)", "Male (%)", "Gap (pp)"),
      digits = 1, booktabs = TRUE,
      caption = "Immigration Intentions by Gender") %>%
  kable_styling(latex_options = c("striped", "scale_down"), font_size = 7)
```

\normalsize

\small
**The gap tells a story:** In `r max_gap_country`, men are `r max_gap_value` percentage points more likely to plan emigration. This likely reflects existing migration networks — men go first, establish themselves, then sometimes bring families.
\normalsize

---

## Wealth Creates Divides

```{r wealth-analysis}
#| echo: false
#| warning: false
#| message: false

# Now I ask: does personal wealth status predict emigration differently across countries?
# First, I calculate emigration rates by wealth status within each country
# CRITICAL: I need to CREATE the wealth_data object here

# Step 1: Check how many people have valid wealth data
# (This ensures we're not accidentally filtering everyone out)
wealth_data <- ab8_youth %>%
  filter(!is.na(wealth_struggling), !is.na(emigrate)) %>%  # Need BOTH variables
  group_by(country_name, wealth_struggling) %>%
  summarise(
    emigrate_pct = weighted.mean(emigrate, w = WT, na.rm = TRUE) * 100,
    n_respondents = n(),
    .groups = "drop"
  ) %>%
  mutate(status = case_when(
    wealth_struggling == 1 ~ "Struggling",
    wealth_struggling == 0 ~ "Comfortable",
    TRUE ~ NA_character_
  ))
```

```{r wealth-gap-calc}
#| echo: false
#| warning: false
#| message: false

# Now I calculate the gap: how much MORE do struggling youth want to leave?
# Reshaping the data to compare Struggling vs Comfortable side-by-side
# A positive gap means struggling youth have higher emigration intentions than comfortable youth
wealth_gap <- wealth_data %>%
  select(country_name, status, emigrate_pct) %>%
  pivot_wider(names_from = status, values_from = emigrate_pct) %>%
  mutate(gap = Struggling - Comfortable) %>%
  arrange(desc(gap))
```

```{r fig-wealth}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 5
#| fig.height: 2.4
#| fig.cap: "Wealth matters differently across countries"

# Now I visualize the relationship between personal wealth and emigration
# I use dodged bars to show Struggling vs Comfortable side-by-side for each country
ggplot(wealth_data, aes(x = country_name, y = emigrate_pct, fill = status)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(
    values = c("Struggling" = "darkred", "Comfortable" = "darkgreen"),
    name = NULL,
    drop = FALSE  # Keep both categories even if one is missing
  ) +
  labs(
    title = "Personal Wealth and Emigration Plans",
    subtitle = "Does being poor make you want to leave?",
    x = NULL, 
    y = "Immigration Intentions (%)"
  ) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.1))) +
  theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(face = "bold", size = 9),
    plot.subtitle = element_text(size = 7),
    legend.position = "bottom",
    legend.text = element_text(size = 7))
```

---

```{r extract-wealth-stats}
#| echo: false
#| warning: false
#| message: false

# Extract the extremes for storytelling in the text
max_wealth_gap_country <- wealth_gap %>% 
  slice_max(gap, n = 1) %>% 
  pull(country_name) %>% 
  as.character()

max_wealth_gap_value <- wealth_gap %>% 
  slice_max(gap, n = 1) %>% 
  pull(gap) %>% 
  round(1)

min_wealth_gap_country <- wealth_gap %>% 
  slice_min(gap, n = 1) %>% 
  pull(country_name) %>% 
  as.character()

min_wealth_gap_value <- wealth_gap %>% 
  slice_min(gap, n = 1) %>% 
  pull(gap) %>% 
  round(1)
```

\small
**Surprising variation:** In `r max_wealth_gap_country`, struggling youth are `r max_wealth_gap_value` percentage points **more likely** to plan emigration than comfortable youth. But in `r min_wealth_gap_country`, the gap is only `r abs(min_wealth_gap_value)` points. **Emigration isn't just about poverty** — it's about a framework of *resources, networks, and opportunity*.
\normalsize

---

## Does National Context Matter?

```{r national-context}
#| echo: false
#| warning: false
#| message: false

# Insight: individuals are nested within countries
# An Iraqi youth and a Jordanian youth might have the same personal wealth,
# but they live in very different national contexts
# Creating country-level variables to capture this

country_context <- ab8_youth %>%
  group_by(country_name) %>%
  summarise(
    nat_econ_fail = weighted.mean(Q101 >= 3, w = WT, na.rm = TRUE) * 100,
    nat_exit_rate = weighted.mean(emigrate, w = WT, na.rm = TRUE) * 100,
    nat_trust_level = weighted.mean(Q301A == 1, w = WT, na.rm = TRUE) * 100
  )

# Now join this back to individual data — each person gets their country's stats
ab8_joined <- ab8_youth %>% left_join(country_context, by = "country_name")

# Extract key comparisons for narrative
lebanon_fail <- country_context %>% filter(country_name == "Lebanon") %>% 
  pull(nat_econ_fail) %>% round(1)
lebanon_exit <- country_context %>% filter(country_name == "Lebanon") %>% 
  pull(nat_exit_rate) %>% round(1)
jordan_fail <- country_context %>% filter(country_name == "Jordan") %>% 
  pull(nat_econ_fail) %>% round(1)
jordan_exit <- country_context %>% filter(country_name == "Jordan") %>% 
  pull(nat_exit_rate) %>% round(1)
```

**The multilevel question:**

Two people, same personal wealth, different countries. One lives where `r lebanon_fail`% say the economy failed (Lebanon). The other where `r jordan_fail`% say so (Jordan). 

Yet Lebanon shows `r lebanon_exit`% exit rate, Jordan shows `r jordan_exit`%. **Same crisis level, different responses. But why?**

---

## Visualizing National Context

```{r fig-context}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 5
#| fig.height: 2.6
#| fig.cap: "National failure doesn't predict exit uniformly"

# I use bubble size to add a third dimension: trust in government
# This helps explain why some failing economies still retain their youth
# Larger bubbles = higher trust, which might keep people from leaving
ggplot(country_context, aes(x = nat_econ_fail, y = nat_exit_rate)) +
  geom_point(aes(size = nat_trust_level), color = "steelblue", alpha = 0.7) +
  geom_text_repel(aes(label = country_name), size = 2.2, fontface = "bold") +
  labs(title = "Economic Failure doesn't mean Automatic Exit",
       x = "National Economic Pessimism (%)",
       y = "Youth Exit Rate (%)") +
  scale_size_continuous(name = "Trust (%)", range = c(3, 9)) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(face = "bold", size = 9),
        legend.position = "bottom",
        legend.text = element_text(size = 7))
```

\small
**Look closely:** No clear line. Tunisia (high failure, high exit). Lebanon (extreme failure, moderate exit). Jordan (high failure, low exit). **Something else is triggering this!**
\normalsize

---

## Testing: Are These Patterns Real?

```{r statistical-tests}
#| echo: false
#| warning: false
#| message: false

# With only 8 countries and noise in survey data, I need to check:
# Could these patterns be random chance?

# Test 1: Do emigration rates truly differ across countries?
# Chi-squared test compares observed vs expected if rates were identical
emigrate_counts <- ab8_youth %>%
  filter(!is.na(emigrate), !is.na(country_name)) %>%
  count(country_name, emigrate) %>%
  pivot_wider(names_from = emigrate, values_from = n, values_fill = 0)
chisq_emigrate <- chisq.test(emigrate_counts[, -1])

# Test 2: Gender gap - is it statistically significant?
# t-test compares means between two groups (male vs female)
gender_test <- t.test(emigrate ~ male, data = ab8_youth)
```

\footnotesize

```{r test-table}
#| echo: false
#| warning: false
#| message: false

# I create a summary table showing test results and what they mean
tibble(
  Test = c("Countries differ on emigration?", "Gender difference in emigration"),
  Result = c(
    paste0("χ² = ", round(chisq_emigrate$statistic, 1), ", p < 0.001"),
    paste0("t = ", round(gender_test$statistic, 2), ", p < 0.001")
  ),
  Interpretation = c("Yes — not random", "Yes — highly significant")
) %>%
  kable(booktabs = TRUE, caption = "Statistical Verification") %>%
  kable_styling(latex_options = "striped", font_size = 7)
```

\normalsize

\small
**Translation:** p < 0.001 means less than 0.1% chance these patterns are random noise. **They're real.**
\normalsize

---

## What Predicts Exit vs Voice?

```{r regression-setup}
#| echo: false
#| warning: false
#| message: false

# Now for the big question: what CAUSES exit vs voice?
# I use logistic regression because my outcomes are binary (yes/no)
# And I use survey-weighted models to respect the sampling design

# Set up survey design object (tells R about the weights and structure)
ab8_design <- svydesign(ids = ~1, strata = ~COUNTRY, weights = ~WT,
                        data = ab8_joined, nest = TRUE)

# Model 1: What predicts EMIGRATION?
# I include: national economy, personal economy, age, gender, urban/rural
model_emigrate <- svyglm(
  emigrate ~ econ_national + Q102 + Q1001 + male + Q13,
  design = ab8_design, family = quasibinomial()
)

# Model 2: What predicts PROTEST?
# Same predictors, different outcome
model_protest <- svyglm(
  protest ~ econ_national + Q102 + Q1001 + male + Q13,
  design = ab8_design, family = quasibinomial()
)

# Extract coefficients and combine for visualization
# I use tidy() from broom package to turn regression output into a clean data frame
coef_both <- bind_rows(
  tidy(model_emigrate, conf.int = TRUE) %>% 
    filter(term != "(Intercept)") %>% 
    mutate(model = "Immigration"),
  tidy(model_protest, conf.int = TRUE) %>% 
    filter(term != "(Intercept)") %>% 
    mutate(model = "Protest")
) %>%
  mutate(term = recode(term,
                       econ_national = "National Economy (worse)",
                       Q102 = "Personal Economy (worse)",
                       Q1001 = "Age",
                       male = "Male",
                       Q13 = "Urban"))
```

```{r fig-regression}
#| echo: false
#| warning: false
#| message: false
#| fig.width: 5.1
#| fig.height: 2.5
#| fig.cap: "Different drivers for exit vs voice"

# I plot coefficients with confidence intervals
# Positive coefficient = increases likelihood, Negative = decreases likelihood
# The confidence intervals show uncertainty - if they cross zero, effect is uncertain
ggplot(coef_both, 
       aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.3, linewidth = 0.8, 
                 position = position_dodge(width = 0.5)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.5)) +
  scale_color_manual(values = c("Immigration" = "orangered", 
                                "Protest" = "dodgerblue"), name = NULL) +
  labs(title = "What Drives Each Response?",
       x = "Effect Size (log-odds)", y = NULL) +
  theme_minimal(base_size = 8) +
  theme(plot.title = element_text(face = "bold", size = 9),
        legend.position = "bottom",
        legend.text = element_text(size = 7))
```

---

## The Key Insight

**Different mechanisms for different choices:**

\medskip

**For EMIGRATION (exit):**

Personal economy matters most. "MY situation is bad" meaning **"I need to leave"**

Please note how the orange (Immigration) bar for personal economy is positive and doesn't cross zero, **this means a strong effect**.

\medskip

**For PROTEST (voice):**

National economy matters most. "OUR country is failing". Simply put, **We need to demand change**

The blue (Protest) bar for national economy shows this collective framing.

\medskip

**Why this matters:**

Exit is an individual escape strategy. Voice is a collective action strategy. **They respond to different triggers.**

---

## Conclusion

### Findings

**Three response patterns:**

1. **High exit, low voice:** Tunisia, Kuwait (individual escape)
2. **Low exit, high voice:** Iraq, Mauritania (collective action)
3. **Paralysis:** Jordan, Palestine (neither option feels viable)

\medskip

**Different mechanisms:**

- Exit responds to **personal** economic hardship
- Voice responds to **national** economic failure
- Gender and wealth create access barriers differently across countries

\medskip

**The bigger story:**

When economies fail, youth aren't passive. But **context determines which option they reach for.**

---

### What I Learned (Beyond Findings)

**Technical skills:**

Survey weighting, multilevel thinking, logistic regression, inline R code for dynamic reporting

\medskip

**Conceptual insights:**

Same economic crisis **is NOT** same response. Individual vs collective framing matters. Missing data requires careful decisions.

\medskip

**Limitations I acknowledge:**

Only 8 countries due to funding. Analysis is cross-sectional (snapshot, not causation). This only represents static data, not a dynamic pattern. Can't account for emigration *costs* (visas, networks, resources, foreign country immigration laws).

---

## Thank You

\bigskip

**Acknowledgments:**

Gemini helped me identify variable codes, troubleshoot survey design setup, and debug code errors. I made all interpretation, visualization, and analytical decisions.

\bigskip

**Methods:** Survey-weighted logistic regression, multilevel analysis via joins, chi-squared and t-tests for verification

**Data:** Arab Barometer Wave VIII (2023-2024) • N = `r format(youth_count, big.mark=",")` youth ages 18-35

---

## References

\footnotesize

Arab Barometer. (2024). *Arab Barometer Wave VIII* [Data set]. [https://www.arabbarometer.org/surveys/arab-barometer-wave-viii/#data_sets](https://www.arabbarometer.org/surveys/arab-barometer-wave-viii/#data_sets)

R Core Team. (2024). *R: A Language and Environment for Statistical Computing*. R Foundation for Statistical Computing.

Wickham, H. et al. (2019). Welcome to the tidyverse. *Journal of Open Source Software*, 4(43), 1686.

Google Gemini (2026). https://gemini.google.com/share/e5dc7d091e9b

\normalsize